---
title: "Analysis of longitudinal data"
date: 09 Dec 2018
output:
html_document:
theme: cerulean
---

<style type="text/css">

body{ /* Normal  */
      font-size: 12px;
  }
td {  /* Table  */
  font-size: 8px;
}
h1.title {
  font-size: 30px;
  color: DarkRed;
}
h1 { /* Header 1 */
  font-size: 22px;
  color: DarkBlue;
}
h2 { /* Header 2 */
    font-size: 20px;
  color: DarkBlue;
}
h3 { /* Header 3 */
  font-size: 18px;
  font-family: "Times New Roman", Times, serif;
  color: DarkBlue;
}
h4 { /* Header 4 */
  font-size: 14px;
  font-family: "Times New Roman", Times, serif;
  color: Black;
}
code.r{ /* Code block */
    font-size: 12px;
}
pre { /* Code block - determines code spacing between lines */
    font-size: 14px;
}
</style>


# Analysis of longitudinal data (Week 6)

## Principle Components Analysis (PCA)

### Read Data
```{r echo=T}
library("readxl")
library("ggplot2")
library("data.table")
library("dplyr")
library("stringr")
library("tidyverse")
library("GGally")
library("ggplot2")
library("MASS")
library("corrplot")
library("plotly")
library("purrr")
library("devtools")
library("ggbiplot")
library("gridExtra")
library("grid")
library("lattice")
library("FactoMineR")

analysis.data <- read.table("~/Documents/GitHub/IODS-project/data/humans2.txt", 
                    header=TRUE)
```


### Description of data
```{r echo=T}
glimpse(analysis.data)
```
