---
title: "Clustering and classification"
date: 25 Nov 2018
output:
html_document:
theme: cerulean
---

<style type="text/css">

body{ /* Normal  */
      font-size: 12px;
  }
td {  /* Table  */
  font-size: 8px;
}
h1.title {
  font-size: 30px;
  color: DarkRed;
}
h1 { /* Header 1 */
  font-size: 22px;
  color: DarkBlue;
}
h2 { /* Header 2 */
    font-size: 20px;
  color: DarkBlue;
}
h3 { /* Header 3 */
  font-size: 18px;
  font-family: "Times New Roman", Times, serif;
  color: DarkBlue;
}
h4 { /* Header 4 */
  font-size: 14px;
  font-family: "Times New Roman", Times, serif;
  color: Black;
}
code.r{ /* Code block */
    font-size: 12px;
}
pre { /* Code block - determines code spacing between lines */
    font-size: 14px;
}
</style>


# Clustering and classification (Week 4)

## Read Data
```{r echo=T}
library("readxl")
library("ggplot2")
library("data.table")
library("dplyr")
library("stringr")
library("tidyr")
library("GGally")
library("ggplot2")
library("MASS")

data("Boston")
```

### Description of data
```{r echo=T}
glimpse(Boston)
```

#### Boston dataset contains the Housing Values in Suburbs of Boston, collected by the U.S Census Service. It is available in the package MASS, and originally derived from the paper: Harrison, D. and Rubinfeld, D.L. (1978) Hedonic prices and the demand for clean air. < em >J. Environ. Economics and Management < b >5, 81–102. 

#### The dataset contains a total of 506 cases with 14 attributes/variables for each case of the dataset.They are:
#### CRIM - per capita crime rate by town
#### ZN - proportion of residential land zoned for lots over 25,000 sq.ft.
#### INDUS - proportion of non-retail business acres per town.
#### CHAS - Charles River dummy variable (1 if tract bounds river; 0 otherwise)
#### NOX - nitric oxides concentration (parts per 10 million)
#### RM - average number of rooms per dwelling
#### AGE - proportion of owner-occupied units built prior to 1940
#### DIS - weighted distances to five Boston employment centres
#### RAD - index of accessibility to radial highways
#### TAX - full-value property-tax rate per $10,000
#### PTRATIO - pupil-teacher ratio by town
#### B - 1000(Bk - 0.63)^2 where Bk is the proportion of blacks by town
#### LSTAT - % lower status of the population
#### MEDV - Median value of owner-occupied homes in $1000's




