---
title: "Regression and model validation"
date: 11 Nov 2018
output:
html_document:
theme: cerulean
---

<style type="text/css">

body{ /* Normal  */
      font-size: 12px;
  }
td {  /* Table  */
  font-size: 8px;
}
h1.title {
  font-size: 30px;
  color: DarkRed;
}
h1 { /* Header 1 */
  font-size: 22px;
  color: DarkBlue;
}
h2 { /* Header 2 */
    font-size: 20px;
  color: DarkBlue;
}
h3 { /* Header 3 */
  font-size: 18px;
  font-family: "Times New Roman", Times, serif;
  color: DarkBlue;
}
code.r{ /* Code block */
    font-size: 12px;
}
pre { /* Code block - determines code spacing between lines */
    font-size: 14px;
}
</style>


## Read Data
```{r echo=T}
analysis_data <- read.csv("~/Documents/GitHub/IODS-project/data/learning2014.csv", 
                    sep=",", header=TRUE)
```

## Description of data
```{r echo=T}
str(analysis_data)
```
The dataset contains a total of 166 observations with 7 variables like gender, age, attitude score and other questions. 

```{r echo=T}
dim(analysis_data)
```
the above command returns the exact dimensions of the dataset we have. The dataset confirms that there are 166 rows with 7 columns containing the variables.  

## Overview of the dataset

### Variable summaries
```{r echo=T}
summary(analysis_data$gender)
```

```{r echo=T}
summary(analysis_data$age)
```

```{r echo=T}
summary(analysis_data$attitude)
```

```{r echo=T}
summary(analysis_data$deep)
```

```{r echo=T}
summary(analysis_data$stra)
```

```{r echo=T}
summary(analysis_data$surf)
```

```{r echo=T}
summary(analysis_data$points)
```

### Load the GGally and ggplot2 libraries
```{r echo=T}
library(GGally)
library(ggplot2)
```

### create plot matrix with ggpairs()
```{r echo=T}
ggpairs(analysis_data, mapping = aes( col = gender, alpha = 0.3 ), lower = list(combo = wrap("facethist", bins = 20)))
```

### 

